Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-09-14 vertical-thermal-review.Rmd
photosynthesis section

**Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies [Table `r table_leaf_metabolism`; @niinemetsPhotosynthesisResourceDistribution2007 ; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].**
This is primarily driven by the greater light available to sun leaves, which also have traits enabling higher photosynthetic rates under high light and sufficient nutrient and water conditions (Tables `r table_leaf_traits`, `r table_leaf_metabolism`).

@@

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Based on consistent positive relationships between growth temperature and optimum temperature of photosynthesis ($T_{opt}$) across sites and seasons [e.g. @tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves [Campbell and Norman 1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent among recent studies, where for three species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019; @hernandezSimilarTemperatureDependence2020]. Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018]. In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021a].

@hernandezSimilarTemperatureDependence2020 also speculated that the observed lack of sun-shade differences in $T_{opt}$ may reflect acclimation to similar temperatures of peak photosynthesis. Sun leaves experience higher temperatures, but maximum temperatures are associated with conditions of midday stomatal depression, and acclimation to optimize photosynthesis at these temperatures would not be advantageous. Shade leaves that do not experience midday depression continue to fix carbon even when afternoon air temperatures peak [@millerOnlySunlitLeaves2021]. Evaluating this requires diurnal monitoring of net photosynthesis through the canopy.

>>>>

Photosynthetic capacity is generally higher in exposed canopy positions-- a fact that is both theoretically expected and observed in numerous field studies [Table `r table_leaf_metabolism`, @niinemetsPhotosynthesisResourceDistribution2007 ; @kenzoHeightrelatedChangesLeaf2015 ; @slotPhotosyntheticHeatTolerance2019; @chenLeafEconomicsSpectrum2020].
This is primarily driven by the greater light available to sun leaves, which also have traits enabling higher photosynthetic rates under conditions of high light and sufficient water (Tables `r table_leaf_traits`, `r table_leaf_metabolism`).

@@

We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves, and existing studies reveal no pronounced overall trend with height in the optimum temperatures for photosynthetic processes (Table `r table_leaf_metabolism`).
Based on consistent positive relationships between growth temperature and optimum temperature of photosynthesis ($T_{opt}$) across sites and seasons [e.g. @tanOptimumAirTemperature2017; @kumarathungeAcclimationAdaptationComponents2019a], one might expect sun leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves [Campbell and Norman 1998; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. 
However, such a trend is not apparent among recent studies, where for three species in Panama, the $T_{opt}$ for sun leaves tended to be moderately, but not significantly, higher than that of shade leaves [@slotPhotosyntheticHeatTolerance2019; @hernandezSimilarTemperatureDependence2020]. 
Similarly, $T_{opt}$ of RuBP carboxylation  ($V_{max}$) and regeneration rates ($J_{max}$) did not differ systematically between tropical sun and shade leaves [@hernandezSimilarTemperatureDependence2020], and along vertical gradients of tropical and temperate trees [@mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021;@mauTemperateTropicalForest2018]. 
In contrast, tropical experimental warming observations showed that $T_{opt}$, and $T_{opt}$ of $J_{max}$ decrease from the understory to the canopy, potentially linked to greater thermal sensitivity of $g_s$ in upper canopy leaves [@carterExperimentalWarmingTropical2021a]. 
@hernandezSimilarTemperatureDependence2020 also speculated that the observed lack of sun-shade differences in $T_{opt}$ may reflect acclimation to similar temperatures of peak photosynthesis. 
Sun leaves experience higher temperatures, but maximum temperatures are associated with conditions of midday stomatal depression, and acclimation to optimize photosynthesis at these temperatures would not be advantageous. 
Shade leaves that do not experience midday depression continue to fix carbon even when afternoon air temperatures peak [@millerOnlySunlitLeaves2021]. 
Evaluating this requires diurnal monitoring of net photosynthesis through the canopy.
